### 2014-11-03 thesis/abstract/abstract.tex
A whole bunch of updates including source code listing in the appendix

Predicting the length of hospital stay for injured patients involves making a
judgement about how long we believe they will need to be hospitalised, so that
doctors and hospital staff can plan the appropriate course of treatment for the
patient to ensure they recover in the quickest possible time. This ensures the
most efficient use of limited hospital resources. However, previous studies
have used a limited number of statistical techniques such as logistic
regression in order to predict the length of stay.

Our work systematically
evaluates the effect of discretisation on all combinations of 11 classifiers
using 5 different feature selection methods. One of these classifiers is a
novel contribution of our work. Furthermore, our evaluation includes 4 feature
selection methods and 3 classifiers which have to date not been used in
predicting the length of stay, compared alongside commonly used
techniques. We also compare two independent data sets.

Our results show that while logistic regression tended to perform best out of
many combinations, it is possible to achieve to achieve an accuracy and an area
under the receiver operating characteristic curve that is less than 2\% worse
than this best result using only around 15\% of the total features. We were
also able to improve upon the baseline accuracy and area under the curve by
2.75\% and 0.034 respectively, and recommend further investigation into the
use of nearest neighbour approaches and feature selection to future studies
of length of stay prediction.

>>>>

Predicting the length of hospital stay for injured patients involves making a
judgement about how long they need to be hospitalised, so that doctors and
hospital staff can plan the appropriate course of treatment for the patient
to ensure they recover in the quickest possible time. This also ensures the
most efficient use of limited hospital resources. However, previous studies
have used a limited number of statistical techniques such as logistic
regression in order to predict the length of stay. Also, they have used only
manual feature selection and have not investigated the effect of automatic
feature selection on classifier performance.

Our work systematically evaluates and compares the performance of a number of
state-of-the-art classification algorithms  and feature selection methods,
and also assesses the effect of feature discretisation for all combinations
of classifiers and feature selectors.  Our evaluation were conducted on two
datasets, one from a trauma ward in a major Sydney hospital and another one
from a general hospital in Portugal. We propose a new classification algorithm
called Ranked Distance Nearest Neighbor that considers the
relative importance of each feature when determining the nearest neighbours.

For the trauma dataset, we were able to achieve an accuracy of 77.81\% with
1-nearest neighbour and an area under the curve of 0.846 with logistic
regression. This represents an improvement of 2.75\% and 0.034 respectively
over the previous best result on the same dataset. For the
general hospital dataset, the best results were 98.23\% accuracy with a
decision tree and 0.994 area under the curve with logistic regression,
support vector machine and K*. We also show that our proposed Ranked
Distance Nearest Neighbour improves accuracy and area under the curve over
traditional nearest neighbour algorithms. We comprehensively discuss
the results and suggest new directions for future studies.

### 2014-11-03 thesis/conclusion/conclusion.tex
A whole bunch of updates including source code listing in the appendix

In this thesis we conducted a systematic and empirical evaluation of a number
of classification algorithms and feature selection methods, as well as the
effect of discretisation, for predicting the length of hospital stay of trauma
patients. In particular, we introduced a new distance function, Ranked
Distance, for the standard
$k$-nearest neighbours algorithm and compared its accuracy and AUC with ten
other classifiers -- Zero-R, 1R, Na\"{i}ve Bayes, C4.5 decision tree, logistic
regression, support vector machine, $k$-nearest neighbour with $k=1$ and 20,
K* and multi-layer perceptron -- several of which (for example K*) have not
been used in existing work on LOS prediction. We also investigated a variety of
feature selection methods -- CFS, C4.5 wrapper, correlation coefficient,
information gain, 1R and expert selection -- encompassing both manual and
automatic methods, many of which have not been applied to the problem of LOS
prediction. Our experiments were conducted on two data sets, one from the
trauma ward of a Sydney hospital and the other a general hospital-wide data
set from Portugal.

\section{Conclusions}
The best accuracy we achieved on the trauma data was 77.81\% with the $k$-NN
classifier using 1 nearest neighbour and 11 out of 78 features. This represents
an improvement over previous work done by Dinh et al. on the same data set,
where they managed to achieve 75.06\% with logistic regression. More
importantly, we were able to improve upon their AUC of 0.812 with a result of
0.846, also using logistic regression but first discretising numeric features
and reducing the feature set to 29 features.
If we consider an even smaller number of features, then K*
gives an AUC of 0.838 with 11 features selected by the C4.5 wrapper, which
improves upon the result of Dinh et al. while using a little more than half
the number of features. The drawback of $k$-NN and K* is that they require all
examples to be stored and then searched during classification, which becomes
storage- and computation-expensive as the data set increases in size.

@@

An important contribution of our work is the use of several nearest neighbour
algorithms for predicting the LOS class of a patient. We found that these
techniques can achieve good accuracy and discriminating ability if a suitable
set of features is found. In addition, we proposed a variation on the standard
nearest neighbours distance function, in order to take the relative importance
of each feature into account when determining the nearest neighbours of an
example. We found that this achieved better results than the other nearest
neighbour algorithms we evaluated, but only when most of the original feature
set was used.

Although we were able to improve on the work of Dinh et al. and confirm the
importance of some features they and others found in predicting the LOS for
trauma patients, we did not find a single classifier or feature selector
that was superior to the others but instead observed different combinations
giving better results than others in certain situations. This highlights the
importance of understanding the data and investigating a range of classifiers
and feature selection techniques when approaching any medical data mining
problem, for which our work will become a catalogue.
However, we did find that logistic regression tended to perform well
on both data sets and across a number of feature selection methods, supporting
its common use in deriving classifiers for various medical phenomena.

@@

For the $k$-NN algorithm using our Ranked Distance approach, we only considered
one kind of weight assignment function, but it is not the only way to assign
these weights. There is room to investigate other types of functions, as well
as using weights obtained from another algorithm (such as logistic regression).
In addition to assigning the weights, we have also only investigated one method
of determining the relative importance of features: the correlation coefficient.
This leaves much room to use other ways of ranking features, such as the
information gain and 1R accuracy (which we have used in our work as feature
selectors).

>>>>

In this thesis we conducted a systematic and empirical evaluation of a number
of classification algorithms and feature selection methods, as well as the
effect of discretisation, for predicting the length of hospital stay of trauma
patients. In particular, we introduced a new classification algorithm, Ranked
Distance Nearest Neighbours, and compared its accuracy and AUC with ten
other classifiers -- Zero-R, 1R, Na\"{i}ve Bayes, C4.5 decision tree, logistic
regression, support vector machine, $k$-nearest neighbour with $k=1$ and 20,
K* and multi-layer perceptron -- several of which (for example K*) have not
been used in existing work on LOS prediction. We also investigated a variety of
feature selection methods -- CFS, C4.5 wrapper, correlation coefficient,
information gain, 1R and expert selection -- encompassing both manual and
automatic methods, many of which have not been applied to the problem of LOS
prediction. Our experiments were conducted on two data sets, one from the
trauma ward of a Sydney hospital and the other a general hospital-wide data
set from Portugal.

\section{Conclusions}
The best accuracy we achieved on the trauma data was 77.81\% with the $k$-NN
classifier using 1 nearest neighbour and 11 out of 78 features. This represents
an improvement over previous work done by Dinh et al. \cite{Dinh2013a}
on the same data set,
where they managed to achieve 75.06\% with logistic regression. More
importantly, we were able to improve upon their AUC of 0.812 with a result of
0.846, also using logistic regression but first discretising numeric features
and reducing the feature set to 29 features.
If we consider an even smaller number of features, then K*
gives an AUC of 0.838 with 11 features selected by the C4.5 wrapper, which
improves upon the result of Dinh et al. \cite{Dinh2013a} while using a little
more than half
the number of features. The drawback of $k$-NN and K* is that they require all
examples to be stored and then searched during classification, which becomes
storage- and computation-expensive as the data set increases in size.

@@

An important contribution of our work is the use of several nearest neighbour
algorithms for predicting the LOS class of a patient. We found that these
techniques can achieve good accuracy and discriminating ability if a suitable
set of features is found. In addition, we proposed a new algorithm, Ranked
Distance Nearest Neighbours, which is an extension of the standard nearest
neighbours classifier that takes the relative importance
of each feature into account when determining the nearest neighbours of an
example. We found that this achieved better results than the other nearest
neighbour algorithms we evaluated, but only when most of the original feature
set was used.

Although we were able to improve on the work of Dinh et al. \cite{Dinh2013a}
and confirm the
importance of some features they and others found in predicting the LOS for
trauma patients, we did not find a single classifier or feature selector
that was superior to the others but instead observed different combinations
giving better results than others in certain situations. This highlights the
importance of understanding the data and investigating a range of classifiers
and feature selection techniques when approaching any medical data mining
problem, for which our work will become a catalogue.
However, we did find that logistic regression tended to perform well
on both data sets and across a number of feature selection methods, supporting
its common use in deriving classifiers for various medical phenomena.

@@

In our proposed Ranked Distance Nearest Neighbours algorithm, we only considered
one kind of weight assignment function, but it is not the only way to assign
these weights. There is room to investigate other types of functions, as well
as using weights obtained from another algorithm (such as logistic regression).
In addition to assigning the weights, we have also only investigated one method
of determining the relative importance of features: the correlation coefficient.
This leaves much room to use other ways of ranking features, such as the
information gain and 1R accuracy (which we have used in our work as feature
selectors).

### 2014-11-03 thesis/introduction/introduction.tex
A whole bunch of updates including source code listing in the appendix

\section{Contributions}
The work presented in this thesis aims to address several gaps in the current
research into LOS prediction, by building upon the work of Dinh et al.
\citep{Dinh2013a}. We systematically and empirically evaluated a comprehensive
catalogue of 11 classification
algorithms and 5 feature selection methods, as well as the effect of
discretisation, for predicting the LOS of trauma patients. 3 of the
classifiers and 4 of the feature selection methods have not been applied in
LOS prediction to date. Additionally, we are also not aware of any previous
work using the discretisation method that we evaluate in this thesis.

We also proposed a new distance function, Ranked Distance, for use with one
of the classification algorithms that we evaluate, and showed that this
performed better than the standard distance function in certain situations.

Finally, we conducted our evaluations on two independent data sets from
different medical domains: one from trauma and the other a general hospital
data set. Not only is the evaluation of a second data set in one work
uncommon in the literature, but the inclusion of more than one data set
allowed us to draw additional insights about the classifiers and feature
selections we used which would not have been possible with only one set of
data.

>>>>

\section{Contributions}
The work presented in this thesis aims to address several gaps in the current
research into LOS prediction, by building upon the work of Dinh et al.
\citep{Dinh2013a}. Our contributions can be summarised as follows:

\begin{enumerate}
\item We systematically evaluated the performance of a catalogue of eleven
classification algorithms and five feature selection methods for predicting the
LOS of trauma patients. Three of the classifiers and four of the feature
selection methods have not been applied in LOS prediction to date, and only
one has been applied in our medical domain.

\item We assessed the effect of feature discretisation on classifier
performance. This has not been studied in previous work.

\item We proposed a new algorithm, Ranked Distance Nearest Neighbour, and
showed that it performed better than the standard nearest neighbour algorithm
in certain situations.

\item We conducted our evaluations on two independent data sets from different
medical domains: one from a trauma ward in a major Sydney hospital and the other
from a general hospital in Portugal. The collection, processing and
understanding of the data involved collaboration with medical experts (Dr
Michael Dinh from the Royal Prince Alfred Hospital in Sydney) and researchers
from Portugal (Associate Professor Paulo Cortez from the University of Minho).
Not only is the evaluation of a second data set in one work uncommon in the
literature, the inclusion of more than one data set allowed us to draw
additional insights about the classifiers and feature selections we used
which would not have been possible with only one set of data.

\item We comprehensively analysed and discussed various aspects of the results
(such as comparison across classification algorithms, feature selection
methods, performance measures, datasets and discretisation). Our results showed
an improvement over the best result on the same datasets.
\end{enumerate}

### 2014-11-03 thesis/method/evaluation.tex
Add example ROC curve to evaluation
A whole bunch of updates including source code listing in the appendix

The methods of evaluating the performance of classifiers in LOS prediction
that are discussed in the literature encompass a wide range of medical
domains and learning problems, which presents us with a challenge when we
decide how to evaluate our work. Here we have chosen to rigorously evaluate
our results (such as using stratified cross-validation when others may have
used a single training-test split), while still using performance measures
comparable to the literature and suitable for our particular learning problem.
This chapter discusses the data sets, procedure and performance measures used
in evaluating our work.

\section{Data Sets}
In our work we obtained two independent sets of LOS data, one specifically for
trauma and the other a general, hospital-wide set.

\subsection{Trauma LOS Data}
\subsubsection{Collection}
The data set used consisted of trauma registry data from the trauma centre
at the Royal Prince Alfred Hospital, a major trauma centre in New South Wales,
Australia. It covered all adult (age 15 and over) inpatient admissions to the
trauma centre from 2007--2011.

All patients were first admitted to the trauma ward until discharged
or transferred to an appropriate unit within the hospital. A single trained
data manager recorded a variety of attributes about the admitted patient,
such as age, gender, blood pressure, mechanism of injury and body regions
that were injured. We received this data in a Microsoft Excel spreadsheet.

\subsubsection{Characteristics}
There were 2546 patient records in the data set we received, comprising of 79
features, one of which was the target variable \texttt{los48}.
\texttt{los48} was a binary variable -- that is, it could only take two
values, 0 or 1 -- with 1 indicating that the patient stayed two days or less,
and 0 indicating a stay of greater than 2 days.

\subsection{General Hospital LOS Data}
By collaborating with our colleagues from the University of Minho in Portugal,
we were able to obtain LOS data for all inpatient hospitalisations from
2000--2013 at the Hospital das Foras Armadas in Lisboa, Portugal. The data
contained 26462 inpatient records spanning all medical departments within the
hospital, comprising 15 features that were not specific to any medical domain.

@@

\subsubsection{Code Written}
\label{sec:code}
\todo{attach code in appendix?}
Leading and trailing whitespace was removed and feature names transformed using
code written in Python to manipulate the CSV. All runs of WEKA were executed by
a program we wrote in Bash (a Unix shell) script. This program was able to run
WEKA with the following features:
\begin{itemize}
\item The ability to specify filenames corresponding to the input training data
and the specification of classification or feature selection algorithms to run
\item ROC curve data output can be optionally enabled and saved to disk after
each cross-validation pass
\item An experiment can be repeated an arbitrary number of times -- this is
specified by the user upon invoking the script (default is one time)
\item The ability to specify the inclusion of a random number seed -- this is
necessary for repetitions of runs to generate different cross-validation splits
\item Model information and performance measures are saved to disk after each
cross-validation
\end{itemize}

@@

In order to implement our novel Ranked Distance function (Section
\ref{sec:rankeddist}), we created our own class within the WEKA Java package
hierarchy. This was implemented in Java and needed to implement the correct
interfaces and override the right inherited methods in order for the $k$-NN
classifier in WEKA to be able to use it as a distance function. This also
involved updating certain properties of the \texttt{weka.jar} executable in
order to use the new class, as well as building the entire WEKA source. The
development environment we used was Eclipse \todo{insert version}, and the
source was built using the 1.7 Java runtime. \todo{double check}

@@

\begin{figure}[h]
\caption{Two ROC curves on the same graph. Note that the straight line from
the bottom left corner to the top right corner indicates an AUC of 0.5. Curves
which head towards the top left corner indicate increasing AUC.}
\label{fig:sampleroc}
\centering
%\includegraphics[scale=0.95]{images/a2-nn}
\end{figure}

>>>>

The methods of evaluating the performance of classifiers in LOS prediction
that are discussed in the literature encompass a wide range of medical
domains and learning problems, which presents us with a challenge when we
decide how to evaluate our work. Here we have chosen to rigorously evaluate
our results (such as using stratified cross-validation when others may have
used a single training-test split), while still using performance measures
comparable to the literature and suitable for our particular learning problem.
This chapter discusses the procedure and performance measures used
in evaluating our work.

@@

\subsubsection{Code Written}
\label{sec:code}
Leading and trailing whitespace was removed and feature names transformed using
code written in Python to manipulate the CSV exported from Microsoft Excel.
All runs of WEKA were executed by
a program we wrote in Bash (a Unix shell) script. This program was able to run
WEKA with the following features:
\begin{itemize}
\item The ability to specify filenames corresponding to the input training data
and the specification of classification or feature selection algorithms to run
\item ROC curve data output can be optionally enabled and saved to disk after
each cross-validation pass
\item An experiment can be repeated an arbitrary number of times -- this is
specified by the user upon invoking the script (default is one time)
\item The ability to specify the inclusion of a random number seed -- this is
necessary for repetitions of runs to generate different cross-validation splits
\item Model information and performance measures are saved to disk after each
cross-validation
\end{itemize}

@@

In order to implement our novel Ranked Distance Nearest Neighbour algorithm,
\ref{sec:rankeddist}), we created our own class within the WEKA Java package
hierarchy. This was implemented in Java and needed to implement the correct
interfaces and override the right inherited methods in order for the $k$-NN
classifier in WEKA to be able to use it as a distance function. This also
involved updating certain properties of the \texttt{weka.jar} executable in
order to use the new class, as well as building the entire WEKA source. The
development environment we used was Eclipse \todo{insert version}, and the
source was built using the 1.7.0 Java runtime.

The source code listing of the most important programs we wrote are displayed
in Appendix \ref{app:code}.

@@

\begin{figure}[h]
\centering
\includegraphics{images/method/example-roc.eps}
\caption{Two ROC curves on the same graph. Note that the straight line from
the bottom left corner to the top right corner indicates an AUC of 0.5. Curves
which head towards the top left corner indicate increasing AUC.}
\label{fig:sampleroc}
\end{figure}

### 2014-11-03 thesis/method/preprocess.tex
A whole bunch of updates including source code listing in the appendix

In this chapter give an overview of preprocessing and describe the steps we
took on our data.

>>>>

In this chapter we introduce the data sets and give an overview of
preprocessing, describing the specific steps we took on our data.

\section{Data Sets}
In our work we obtained two independent sets of LOS data, one specifically for
trauma and the other a general, hospital-wide set.

\subsection{Trauma LOS Data}
\subsubsection{Collection}
The data set used consisted of trauma registry data from the trauma centre
at the Royal Prince Alfred Hospital, a major trauma centre in New South Wales,
Australia. It covered all adult (age 15 and over) inpatient admissions to the
trauma centre from 2007--2011.

All patients were first admitted to the trauma ward until discharged
or transferred to an appropriate unit within the hospital. A single trained
data manager recorded a variety of attributes about the admitted patient,
such as age, gender, blood pressure, mechanism of injury and body regions
that were injured. We received this data in a Microsoft Excel spreadsheet.

\subsubsection{Characteristics}
There were 2546 patient records in the data set we received, comprising of 79
features, one of which was the target variable \texttt{los48}.
\texttt{los48} was a binary variable -- that is, it could only take two
values, 0 or 1 -- with 1 indicating that the patient stayed two days or less,
and 0 indicating a stay of greater than 2 days.

\subsubsection{Features}

\subsection{General Hospital LOS Data}
By collaborating with our colleagues from the University of Minho in Portugal,
we were able to obtain LOS data for all inpatient hospitalisations from
2000--2013 at the Hospital das Foras Armadas in Lisboa, Portugal. The data
contained 26462 inpatient records spanning all medical departments within the
hospital, comprising 15 features that were not specific to any medical domain.

\subsubsection{Features}

### 2014-11-03 thesis/thanks/thanks.tex
A whole bunch of updates including source code listing in the appendix

Thank you to all of my friends, in particular Daniel Buckmaster, Stephen
Caraher and Christine Au Yeung, for the selfless encouragement, help and good
times which have made made this year-long journey possible.

>>>>

Thank you to all of my friends, in particular Daniel Buckmaster, Stephen
Caraher and Christine Au Yeung, for the selfless encouragement, help and good
times which have made this year-long journey possible.
